Pushdown Control-Flow Analysis via Refunctionalization

### 2018-02-24 paper.tex
misc
update

Tranditionally, continuation-passing style is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use a direct-style intermediate representation as target for
clarity but without losing simplicty and generality.
The transformations we will show in the rest of this paper
also work on abstract machines for plain $\lambda$-calculus.

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus. (TODO: cite)
CESK machine models program execution as state transitions in small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only catagory of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

@@

\subsection{Abstracting Abstract Machine} \label{aam}

store allocated continuation
binding store, continuation store
store maps from addresses to a set of values
store update is a join
the step function is non-deterministic

@@

type Time = List[Expr]

case class BAddr(x: String, time: Time)
type Env = Map[String, BAddr]

abstract class Storable 
case class Clos(v: Lam, env: Env) extends Storable
type BStore = Store[BAddr, Storable]

@@

case class Frame(x: String, e: Expr, env: Env)
case class Cont(frame: Frame, kaddr: KAddr)
type KStore = Store[KAddr, Cont]

case class State(e: Expr, env: Env, bstore: BStore, 
                 kstore: KStore, k: KAddr, time: Time)
\end{verbatim}

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, tgtEnv: Env, tgtStore: BStore, time: Time): KAddr = 
  ContAddr(tgtExpr, time)
def k: Int = 0
def tick(s: State): Time = (s.e::s.time).take(k)

@@

\begin{verbatim}
def step(s: State): List[State] = {
  val newTime = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Clos(Lam(v, body), c_env) <- atomicEval(f, env, bstore).toList) yield {
        val baddr = allocBind(v, newTime)
        val newEnv = c_env + (v -> baddr)
        val newBStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val newKAddr = allocKont(body, c_env, newBStore, newTime)                                                          
        val newKStore = kstore.update(newKAddr, Cont(Frame(x, e, env), kaddr))
        State(body, newEnv, newBStore, newKStore, newKAddr, newTime)
      }
    case State(ae, env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Cont(Frame(x, e, f_env), f_kaddr) <- kstore(kaddr).toList) yield {
        val baddr = allocBind(x, newTime)
        val newEnv = f_env + (x -> baddr)
        val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, newEnv, newStore, kstore, f_kaddr, newTime)
      }
  }
}
\end{verbatim}

Comparing with concrete CESK machine, the
\texttt{drive} function performs a collecting semantics rather than valuation
semantics. That is, for the purpose of analyzing a program, \texttt{drive}
collects all the intermediate machine states as program is abstractly executing.
This is a variant of work list algorithm that always applies \texttt{step}
function to the first element of \texttt{todo} and adds it to \texttt{seen}
if there is an unexplored one. If the work list is empty, the function just
returns the set of explored states up to now.

@@

def step(s: State): List[State] = {
  val newTime = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) =>
      for (Clos(Lam(v, body), c_env) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, newTime)
        val newEnv = c_env + (v -> baddr)
        val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, newEnv, newStore, frame::konts, newTime)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, f_env)::konts =>
          val baddr = allocBind(x, newTime)
          val newEnv = f_env + (x -> baddr)
          val newStore = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, newEnv, newStore, konts, newTime))
      }
  }
}

>>>>

Tranditionally, continuation-passing style is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use a direct-style intermediate representation as target for
clarity but without losing simplicty and generality.
The transformations we will show in the rest of this paper
also work on abstract machines for plain $\lambda$-calculus languages.

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
CESK machine models program execution as state transitions in small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only catagory of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine is a systematic methodology that derives abstract interpreters
from a concrete abstract machine. TODO

There are several notable differences of store between AAM and concrete CESK machine.
In AAM, the store maps addresses to sets of values, so it stores all possible values
for a particular address. Thus the dereference an address becomes non-deterministic.
Also, the store performs a join rather than overwriting when updating elements of the store.
Third, the continuations are allocated on store instead of formed a linked list.

For clarity, we divide the store into binding store \texttt{BStore} and continuation store \texttt{KStore}. 
The binding store maps addresses of binding to sets of closure values, and continuation store maps
continuation addresses to sets of continuations.
We define a generic class \texttt{Store[K,V]} that performs join when updating elements
in a store. By parametering \texttt{Store[K,V]} with \texttt{[BAddr, Storable]} and 
\texttt{[KAddr, Cont]}, we obtain \texttt{BStore} and \texttt{KStore}, perspectively.

Note that the store is monotonic. TODO

@@

type BStore = Store[BAddr, Storable]
type KStore = Store[KAddr, Cont]
\end{verbatim}

As a consequence, the components of a state is also changed: 
the store is divided into binding store and continuation store;
the continuation stack becomes an address that maps to a set of continuation.
By dereferencing this address in continuation store, we can retrieve the 
actually transfer of controls.

\begin{verbatim}
case class State(e: Expr, env: Env, bstore: BStore, 
                 kstore: KStore, k: KAddr, time: Time)
\end{verbatim}

The definitions of closure \texttt{Clos}, stack frame \texttt{Frame}, and 
environment \texttt{Env} remain the same as concrete CESK machine.

\begin{verbatim}
case class BAddr(x: String, time: Time)
type Env = Map[String, BAddr]

@@

case class Frame(x: String, e: Expr, env: Env)
case class Cont(frame: Frame, kaddr: KAddr)
\end{verbatim}

To allocate addresses for bindings or continuations, we use \texttt{allocBind}
and \texttt{allocKont}.

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, tgtEnv: Env, tgtStore: BStore, time: Time): KAddr = 
  ContAddr(tgtExpr, time)
type Time = List[Expr]
def k: Int = 0
def tick(s: State): Time = (s.e::s.time).take(k)

@@

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val newBStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val newKAddr = allocKont(body, env_c, newBStore, new_time)                                                          
        val newKStore = kstore.update(newKAddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, newBStore, newKStore, newKAddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Cont(Frame(x, e, f_env), f_kaddr) <- kstore(kaddr).toList) yield {
        val baddr = allocBind(x, new_time)
        val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

Comparing with concrete CESK machine, the
\texttt{drive} function performs a collecting semantics rather than valuation
semantics. That is, for the purpose of analyzing a program, \texttt{drive}
collects all the intermediate machine states as program is abstractly executing.
The following code shows a variant of work list algorithm that always applies \texttt{step}
function to the first element of \texttt{todo} and adds it to \texttt{seen} set
if there is an unexplored one. If the work list is empty, the function just
returns the set of explored states up to now.

@@

def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, f_env)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = f_env + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
